Queue Aware Precoder Design for Space Frequency Resource Allocation

### 2014-07-03 Report/joint_wsrm_formulation.tex
Updated on 16.47:Jul-2
Satya corrections !

The \ac{WSRM} technique is predominantly used to design the precoders for the \ac{MIMO} \ac{BC} channels to maximize the weighted sum throughput in the downlink direction. The weights are used to provide different priorities to the users based on their requirements and \ac{QoS} criterion. Since the objective of this work is mainly concerned with the reduction of the number of queued packets, we use the length of the queued packets for each user as the respective weights. Using the length of the queued packets as the corresponding weights is obtained from the well known algorithm, called the \textit{backpreassure algorithm}, which is used to decide the routing decisions for each node in the network to efficiently send the packets to the desired destination with minimal packet drops and delay. The backpreassure algorithm is the outcome of minimizing the Lyapunov drift conditioned over the current queue states \me{\mbf{Q}(i - 1)}, where \me{\mbf{Q}} is the vector formed by stacking the number of queued packets of each user in the system. Assuming the queue length grows according to \eqref{eqn-2a} for all the users, the Lyapunov drift conditioned over different channel realizations is given by
\begin{equation} \label{eqn-3.1.0}
\Delta(i) = \frac{1}{2} \, \sum_{k \in \mc{U}} \, \mathrm{E} \Big{[}  \left ( Q_k(i)^2 - Q_k(i-1)^2 \right ) \vert \, \mbf{Q}(i - 1) \Big{]},
\end{equation}
where the expectation is taken over all possible control decisions made in accordance to the channel under consideration to minimize the total backlogged packets at a given instant. Now, by substituting \eqref{eqn-2a} in \eqref{eqn-3.1.0}, the resulting expression is given by
\begin{equation}
\Delta(i) \leq \mathrm{E} \Big{[} \mbf{Q}(i-1) \left \lbrace \mbfa{\lambda}(i) - \mbf{t}(i) \right \rbrace | \mbf{Q}(i-1) \Big{]} + \frac{1}{2} \,\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \frac{1}{2} \,\mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ],
\label{drift-exp}
\end{equation}
where \me{\mbfa{\lambda}} and \me{\mbf{t}} are the vectors formed by stacking the arrival and transmission of all users in the system and the inequality is due to the upper bound\footnote{\me{\left ( \max{\Big [ \mbf{Q}(i-1) - \mbf{t}(i),0 \Big ] } + \mbfa{\lambda}(i) \right )^2 \leq \mbf{Q}^2(i-1) + \mbf{t}^2(i) + \mbfa{\lambda}^2(i) + 2 \, \mbf{Q}(i-1) \left ( \mbfa{\lambda}(i) - \mbf{t}(i) \right ) }}.

The first term in \eqref{drift-exp} can be neglected from the optimization, since the arrivals are independent of the current queue state following \me{\mathrm{E} [ \mbf{Q}(i-1) \mbfa{\lambda}(i) | \mbf{Q}(i-1)] = \mbf{Q}(i-1)\mathrm{E} [\mbfa{\lambda}(i)] = \mbf{Q}(i-1)\mbf{A}(i)}. In the backpreassure algorithm, discussed in \cite{georgiadis2006resource,neely2010stochastic}, the second order terms are bounded by a constant \me{B}, \textit{i.e}, \me{\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ] \leq B}. With this approximation, the resulting problem can be represented as
\begin{equation}
\underset{\mbf{t}(i)}{\text{minimize}} \sum_{k \in \mc{U}} \, Q_k(i-1) \: t_k(i).
\label{backpreassure}
\end{equation}
In \eqref{backpreassure}, the expectation operator is dropped since it is maximized by maximizing the function inside it. Now from \eqref{backpreassure}, the optimization variables \me{\mbf{t}(i)}, which is the vectorized transmission rates, are to be identified for each user for the current slot \me{i}. Since the current system can exploit both space and frequency resources through \ac{MIMO} \ac{OFDM} model, the transmission rates \me{\mbf{t}(i)} for the current instant can be controlled using the transmit precoders \me{\mbf{M}_{k,n}} designed for each user over each sub-channel resource. Since the decision variables \me{\mbf{t}(i)} depends on the queue state at \me{\mbf{Q}(i - 1)}, we drop the time variable \me{i} for the easy representation. The above problem defined in \eqref{backpreassure} can be rewritten to include the space-frequency allocation variable, namely \me{\mbf{M}_{k,n}}, as
\begin{IEEEeqnarray}{rCl} \label{q_gen_sum}
\underset{\substack{\mvec{M}{k,n},\gamma_{l,k,n} \\ t_{l,k,n}}}{\text{maximize}} &\quad& \sum_{k \in \mc{U}} \, Q_k \left ( \sum_{n = 1}^N \sum_{l=1}^L t_{l,k,n} \right ) \IEEEyessubnumber \label{eqn-3.1.1} \\
\text{subject to.} & \quad & t_{l,k,n} \leq \log_2(1+\gamma_{l,k,n}) \IEEEyessubnumber \label{eqn-3.1.2} \\
& \quad &\sum_{n = 1}^N \sum_{k \in \mathcal{U}_b} \text{tr} \, (\mvec{M}{k,n} \mvec{M}{k,n}^\herm) \leq P_{{\max}}, \fall b, \IEEEyessubnumber \label{eqn-3.1.3} \\
& \quad & \sum_{n=1}^N \sum_{l = 1}^L t_{l,k,n} \leq Q_k, \fall k \in \mc{U} \IEEEyessubnumber \label{eqn-3.1.4}
\end{IEEEeqnarray}
where \eqref{eqn-3.1.4} is due to the \me{[x]^+} operation in \eqref{eqn-2a} and \me{\gamma_{l,k,n}} is defined in \eqref{eq:SINR}. By using the number of queued packets as the weights, the resources can be allocated to the user with the more number of backlogged packets, which essentially does the allocation in a greedy manner. As a special case of the problem defined in \eqref{q_gen_sum}, we can formulate the sum rate maximization problem by setting the weights in \eqref{eqn-3.1.1} as unity, leading to the problem as
\begin{IEEEeqnarray}{rCl}\label{gen_sum}
\underset{\substack{\mvec{M}{k,n},\gamma_{l,k,n} \\ t_{l,k,n}}}{\text{maximize}} &\quad& \sum_{k \in \mc{U}} \, \sum_{n = 1}^N \sum_{l=1}^L t_{l,k,n} \IEEEyessubnumber \\
\text{subject to.} & \quad & \eqref{eqn-3.1.1}, \eqref{eqn-3.1.2}, \eqref{eqn-3.1.3} \; \text{and} \; \eqref{eqn-3.1.4}. \IEEEyessubnumber
\end{IEEEeqnarray}

>>>>

The \ac{WSRM} technique is predominantly used to design the precoders for the \ac{MIMO} \ac{BC} channels to maximize the weighted sum throughput in the downlink direction. The weights are used to provide different priorities to the users based on the respective \ac{QoS} criterion. Since the objective of this work is concerned with the reduction of the number of queued packets, we use the total number of the queued packets for each user as the respective weights. Using the size of the queued packets as the corresponding weights is obtained from the well known algorithm, called the \textit{backpreassure algorithm}, which is used to decide the routing decisions for each node in the network to efficiently send the packets to the desired destination with minimal packet drops and delay. The backpreassure algorithm is the outcome of minimizing the Lyapunov drift \cite{neely2010stochastic} conditioned over the current queue states \me{\mbf{Q}(i - 1)}, where \me{\mbf{Q}} is the vector formed by stacking the number of queued packets of each user in the system. Assuming the queue length grows according to \eqref{eqn-2a} for all the users, the Lyapunov drift conditioned over the current queue state is given by
\begin{equation} \label{eqn-3.1.0}
\Delta(i) = \frac{1}{2} \, \sum_{k \in \mc{U}} \, \mathrm{E} \Big{[}  \left ( Q_k(i)^2 - Q_k(i-1)^2 \right ) \vert \, \mbf{Q}(i - 1) \Big{]},
\end{equation}
where the expectation is taken over all possible control\footnote{transmission rates} decisions made in accordance to the channel under consideration to minimize the total backlogged packets at a given instant. Now, by substituting \eqref{eqn-2a} in \eqref{eqn-3.1.0}, the resulting expression is given by
\begin{equation}
\Delta(i) \leq \mathrm{E} \Big{[} \mbf{Q}(i-1) \left \lbrace \mbfa{\lambda}(i) - \mbf{t}(i) \right \rbrace | \mbf{Q}(i-1) \Big{]} + \frac{1}{2} \,\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \frac{1}{2} \,\mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ],
\label{drift-exp}
\end{equation}
where \me{\mbfa{\lambda}} and \me{\mbf{t}} are the vectors formed by stacking the arrival and transmission of all users in the system and the inequality provides the upper bound\footnote{\me{\left ( \max{\Big [ \mbf{Q}(i-1) - \mbf{t}(i),0 \Big ] } + \mbfa{\lambda}(i) \right )^2 \leq \mbf{Q}^2(i-1) + \mbf{t}^2(i) + \mbfa{\lambda}^2(i) + 2 \, \mbf{Q}(i-1) \left ( \mbfa{\lambda}(i) - \mbf{t}(i) \right ) }} for the Lyapunov drift.

The first term in \eqref{drift-exp} can be neglected from the optimization, since the arrivals are independent of the current queue state following \me{\mathrm{E} [ \mbf{Q}(i-1) \mbfa{\lambda}(i) | \mbf{Q}(i-1)] = \mbf{Q}(i-1)\mathrm{E} [\mbfa{\lambda}(i)] = \mbf{Q}(i-1)\mbf{A}(i)}. In the backpreassure algorithm, discussed in \cite{georgiadis2006resource,neely2010stochastic}, the second order terms are bounded by a constant \me{B}, \textit{i.e}, \me{\mathrm{E} \left [ \mbfa{\lambda}^2(i) | \mbf{Q}(i-1) \right ] + \mathrm{E} \left [ \mbf{t}^2(i) | \mbf{Q}(i-1) \right ] \leq B}. With this approximation, the resulting problem can be represented as
\begin{equation}
\underset{\mbf{t}(i)}{\text{minimize}} \sum_{k \in \mc{U}} \, Q_k(i-1) \: t_k(i).
\label{backpreassure}
\end{equation}
In \eqref{backpreassure}, the expectation operator is dropped since it is maximized by maximizing the function inside it. Now from \eqref{backpreassure}, the optimization variables \me{\mbf{t}(i)}, which is the vectorized transmission rates, are to be identified for each user for the current slot \me{i}. Since the current system can exploit both space and frequency resources through \ac{MIMO} \ac{OFDM} model, the transmission rates \me{\mbf{t}(i)} for the current instant can be controlled using the transmit precoders \me{\mbf{M}_{k,n}} designed for each user over each sub-channel resource. Since the decision variables \me{\mbf{t}(i)} depends on the queue state at \me{\mbf{Q}(i - 1)}, we drop the time variable \me{i} for the easy representation. The above problem defined in \eqref{backpreassure} can be rewritten to include the space-frequency allocation variable, namely \me{\mbf{M}_{k,n}}, as
\begin{IEEEeqnarray}{rCl} \label{q_gen_sum}
\underset{\substack{\mvec{M}{k,n},\gamma_{l,k,n} \\ t_{l,k,n},\mvec{W}{k,n}}}{\text{maximize}} &\quad& \sum_{k \in \mc{U}} \, Q_k \left ( \sum_{n = 1}^N \sum_{l=1}^L t_{l,k,n} \right ) \IEEEyessubnumber \label{eqn-3.1.1} \\
\text{subject to.} & \quad & t_{l,k,n} \leq \log_2(1+\gamma_{l,k,n}) \IEEEyessubnumber \label{eqn-3.1.2} \\
& \quad &\sum_{n = 1}^N \sum_{k \in \mathcal{U}_b} \text{tr} \, (\mvec{M}{k,n} \mvec{M}{k,n}^\herm) \leq P_{{\max}}, \fall b, \IEEEyessubnumber \label{eqn-3.1.3} \\
& \quad & \sum_{n=1}^N \sum_{l = 1}^L t_{l,k,n} \leq Q_k, \fall k \in \mc{U} \IEEEyessubnumber \label{eqn-3.1.4}
\end{IEEEeqnarray}
where \eqref{eqn-3.1.4} is due to the \me{[x]^+} operation in \eqref{eqn-2a} and \me{\gamma_{l,k,n}} is defined in \eqref{eq:SINR}. By using the number of queued packets as the weights, the resources can be allocated to the user with the more number of backlogged packets, which essentially does the allocation in a greedy manner. As a special case of the problem defined in \eqref{q_gen_sum}, we can formulate the sum rate maximization problem by setting the weights in \eqref{eqn-3.1.1} as unity, leading to the problem as
\begin{IEEEeqnarray}{rCl}\label{gen_sum}
\underset{\substack{\mvec{M}{k,n},\gamma_{l,k,n} \\ t_{l,k,n},\mvec{W}{k,n}}}{\text{maximize}} &\quad& \sum_{k \in \mc{U}} \, \sum_{n = 1}^N \sum_{l=1}^L t_{l,k,n} \IEEEyessubnumber \\
\text{subject to.} & \quad & \eqref{eqn-3.1.1}, \eqref{eqn-3.1.2}, \eqref{eqn-3.1.3} \; \text{and} \; \eqref{eqn-3.1.4}. \IEEEyessubnumber
\end{IEEEeqnarray}

### 2014-07-03 Report/problem_formulation_intro.tex
Satya corrections !

In this paper, we address the problem of resource allocation across the \ac{MIMO} \ac{OFDM} framework, which are shared by the users served by multiple \acp{BS}. In order to minimize the queued packets, the precoders are to be designed in a way to overcome the problem of allocating the resources more than required. In order to satisfy this constraint, we use the queue deviation as the utility function as compared to the traditional \ac{WSRM} objective discussed extensively in the literature \cite{christensen2008weighted,wmmse_shi}. For practical and tractability reasons, we impose a constraint that the maximum number of transmitted bits for the user \me{k} is limited by the packets available at the transmitter. As a result, the number of backlogged packets \me{v_k} remaining in the system for the user \me{k} is given by
\begin{equation}
v_k =  Q_k - \sum_{n = 1}^N \sum_{l = 1}^{L} \log_2(1+\gamma_{l,k,n}) \geq 0 \fall k \in \mathcal{U},
\label{eqn-4.2}
\end{equation}
where the expression \me{t_{l,k,n}} is replaced by maximum achievable rate by using the Gaussian signaling, which is equal to \me{\log_2(1 + \gamma_{l,k,n})} for a given \ac{SINR}. The above positivity constraint need to be satisified by \me{v_k} to avoid the excessive allocation of the resources.

>>>>

In this paper, we address the problem of resource allocation across the \ac{MIMO} \ac{OFDM} framework, which are shared by the users served by multiple \acp{BS}. In order to minimize the queued packets, the precoders are to be designed in a way to overcome the problem of allocating the resources more than required. In order to satisfy this constraint, we use the queue deviation as the utility function as compared to the traditional \ac{WSRM} objective discussed extensively in the literature \cite{christensen2008weighted,wmmse_shi}. For practical and tractability reasons, we impose a constraint that the maximum number of transmitted bits for the user \me{k} is limited by the packets available at the transmitter. As a result, the number of backlogged packets \me{v_k} remaining in the system for the user \me{k} is given by
\begin{equation}
v_k =  Q_k - \sum_{n = 1}^N \sum_{l = 1}^{L} \log_2(1+\gamma_{l,k,n}) \geq 0 \fall k \in \mathcal{U}.
\label{eqn-4.2}
\end{equation}
The above positivity constraint need to be satisified by \me{v_k} to avoid the excessive allocation of the resources.

### 2014-07-03 Report/system_model.tex
Updated ver_1
Satya corrections !

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} for transmission.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received signal of the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n^\mathrm{th}} at user $k$ is given by
\begin{equation}\label{eqn-1}
y_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} with rank \me{L = \min(N_R,N_T)} is the channel between the \ac{BS} \me{b} and  user \me{k} on the sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. Assuming  \me{||\mvec{w}{l,k,n}||_{2}^{2} = 1} and independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}.
\end{equation}
With the infinite buffer model assumption, let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{i}^\mathrm{th}} instant for the user \me{k} depend on the fresh arrivals at the \me{{i}^\mathrm{th}} instant and the total number of backlogged packets \me{Q_k(i-1)} as given by
\begin{equation}
Q_k(i) = \Big [ Q_k(i-1) - t_k(i-1) \Big ]^+ + \lambda_k(i),
\label{eqn-2a}
\end{equation}
where \me{t_k} denote the transmissions in bits and \me{[x]^+ = \max{ \lbrace x,0  \rbrace}}. The total number of transmitted bits for the user \me{k} is given by \me{t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i)},
where \me{t_{l,k,n}} is the total number of bits that can be transmitted over \me{l^\mathrm{th}} spatial stream and on \me{n^\mathrm{th}} sub-channel at the \me{i^\mathrm{th}} instant. Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.

>>>>

In this paper we adopt linear beamforming technique at \acp{BS}. Specifically, the data symbols \me{d_{l,k,n}} for  user \me{k} on the \me{l^\mathrm{th}} spatial stream over the sub-channel \me{n} is multiplied with the beamformer \me{\mvec{m}{l,k,n} \in \mathbb{C}^{N_T \times 1}} for transmission.   In order to detect  multiple spatial streams at the receiver, a receive beamforming vector \me{\mvec{w}{l,k,n}} is employed at each user. Consequently, the received data symbol corresponding to the \me{l^\mathrm{th}} spatial stream over sub-channel \me{n^\mathrm{th}} at user $k$ is given by
\begin{equation}\label{eqn-1}
\hat{d}_{l,k,n} = \mvec{w}{l,k,n}^\herm \mvec{H}{b_k,k,n} \,\mvec{m}{l,k,n} d_{l,k,n} + \mvec{w}{l,k,n}^\herm \sum_{i \in \mc{U} \backslash \set{k}} \mvec{H}{b_i,k,n} \sum_{j = 1}^L \mvec{m}{j,i,n}d_{j,i,n} + \mvec{w}{l,k,n}^\herm \mvec{n}{k,n}
\end{equation}
where \me{\mvec{H}{b,k,n} \in \mathbb{C}^{N_R \times N_T}} with rank \me{L = \min(N_R,N_T)} is the channel between the \ac{BS} \me{b} and  user \me{k} on the sub-channel \me{n}, and \me{\mvec{n}{k,n}} \me{\sim \mc{CN}(0,N_0)} is the additive noise vector for the user \me{k} on the \me{n^\mathrm{th}} sub-channel and \me{l^\mathrm{th}} spatial stream. Assuming  \me{||\mvec{w}{l,k,n}||_{2}^{2} = 1} and independent detection of data streams, we can write the signal-to-interference-plus-noise ratio (SINR) as
\begin{equation}\label{eq:SINR}
\gamma_{l,k,n} = \dfrac{\left |\mvec{w}{l,k,n}^\herm \, \mvec{H}{b_k,k,n} \, \mvec{m}{l,k,n} \right |^2}{N_0 + \sum_{(j,i) \neq (l,k)} |\mvec{w}{l,k,n}^\herm \mvec{H}{b_i,k,n} \mvec{m}{j,i,n} |^2}.
\end{equation}
With the infinite buffer model assumption, let \me{Q_k} be the number of backlogged packets which are destined for the user \me{k} at a given scheduling instant. The queue dynamics of the user \me{k} are modeled using the Poisson arrival process with the average packet arrivals of \me{A_k = \mathbf{E}_i\{\lambda_k\}} packets/bits, where \me{\lambda_k(i) \sim \mathrm{Pois}(A_k)} represents the instantaneous number of packets or bits arriving for the user \me{k} at the \me{{i}^\mathrm{th}} instant. The total number of queued packets at the \me{{i}^\mathrm{th}} instant for the user \me{k} depend on the fresh arrivals at the \me{{i}^\mathrm{th}} instant and the total number of backlogged packets \me{Q_k(i-1)} as given by
\begin{equation}
Q_k(i) = \Big [ Q_k(i-1) - t_k(i-1) \Big ]^+ + \lambda_k(i),
\label{eqn-2a}
\end{equation}
where \me{[x]^+ = \max{ \lbrace x,0  \rbrace}} and \me{t_k} denotes the transmission in bits for user \me{k}. For \ac{MIMO} \ac{OFDM} system,
\begin{equation}
t_k(i) = \sum_{n = 1}^N \, \sum_{l = 1}^L \, t_{l,k,n}(i),
\end{equation}
where \me{t_{l,k,n} = \log_2(1 + \gamma_{l,k,n})} corresponds to the maximum rate achieved by Gaussian signaling on \me{l^\mathrm{th}} spatial stream and over \me{n^\mathrm{th}} sub-channel for the \ac{SINR} of \me{\gamma_{l,k,n}}. Note that the units of \me{t_k} and \me{Q_k} are in bits defined per channel use.
